---
title: "SSC442_Lab_1_E2"
output:
  pdf_document: 
    latex_engine: lualatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

```{r, echo=FALSE, include=FALSE}
library(readr)
library(tidyverse)
library(ggplot2)
library(scales)

urlfile = 'https://raw.githubusercontent.com/MSUDataAnalytics/SSC442/master/Labs/data/bank.csv'
bank = read_csv(url(urlfile))
head(bank)

sub_data = bank[bank$y=='yes',]
```

# Bank Analysis
## Introduction and Methodology

  This memo focuses on two subsets of banking data that help to capture the makeup of those who did and did not choose to subscribe to the long-term savings program. By subsetting our data we're able to find the differences in those who opt in and those who don't which will further allow for better marketing in the future. The analysis is performed using a range of variables that show some correlation. These are balance and job type colored by education and balance and campaign calls colored by previous marketing.

## Balance and Education
```{r, echo=FALSE, out.width='50%'}
S_balance_education = ggplot(data=sub_data,
                          mapping=aes(x=job, y=balance, color = education))

A_balance_education = ggplot(data=bank,
                          mapping=aes(x=job, y=balance, color = education))

s_plot_b_e = S_balance_education + geom_jitter() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

a_plot_b_e = A_balance_education + geom_jitter() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

s_plot_b_e

a_plot_b_e
```

  Above we have a plot showing balance by job type colored by education. The data being fed into this plot is of those who opted in to the product. We highlight the density of the blue-collar, management, and technician customers. By more heavily targetting these groups there should be an increase in subscribers to the surface. The following graph shows a relatively homogenous spread of characteristics between jobs, meaning that our prediction of increased subscribers from these groups should hold when scaled up.


## Campaign and Balance
```{r, echo=FALSE, out.width='50%'}
S_balance_duration = ggplot(data=sub_data,
                          mapping = aes(x=campaign, y=balance, color=previous))

A_balance_duration = ggplot(data=bank,
                          mapping = aes(x=campaign, y=balance, color=previous))

s_plot_b_d = S_balance_duration + geom_point()

a_plot_b_d = A_balance_duration + geom_point()

s_plot_b_d

a_plot_b_d
```

  The above plot shows the patterns derived from campaign efforts and balance. You'll notice by comparison that those who opted in typically have far lower balances than those who did not. It also shows that the marketing efforts typically have little effect after the first 3 attempts and no effect after the first 10 attempts. Our recomendation is to taret those with lower balances and to respect the customers answer after 3-5 attempts. This should help to reduce the time and money spent marketing to customers that have made up there mind.
  
## Conclusion

  Our memo shows correlations between some interesting variables focusing on the type of customer who opts in and the campaigning effectiveness. We believe implementing the suggestion above will help to better target customers who are interested in the product and save money on over-marketing to those who have made up their mind on the product. With more data on the marketing timing we may be able to gain insight into the best time to market to customers, however this data set was able to provide a strong base for changes.